### 2019-08-20 ThesisJo/2_background.tex
2019-08-19

Waves are made of endless tiny energy transactions, a dance of particles effecting each other in a search for equilibrium. They are caused by disturbances in the forces of nature. The drop of a rock into flat water or rapid displacement of still air cause mechanical waves, while synchronized oscillations of electric and magnetic fields cause electromagnetic waves. Regardless of whether the waves are mechanical or electromagnetic, move perpendicular or parallel to the axis of propagation, what medium they propagate through or the means by which they do so, they are---perhaps surprisingly---all governed by a partial differential equation known as the \emph{wave equation}:
%
\begin{align}
\Delta f = 
\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2} = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
\end{align}
%
where $\Delta$ is the Laplace operator, $f$ is field strength and $c\in\mathbb{R}^+$ is wave propagation speed. If $f$ represents the displacement of a vibrating string, the equation states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side). In acoustics $f$ represents sound pressure, but since the mathematics are all the same we will simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.

Suppose an array of sensors has a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length). A signal source is located at a point $\udot{S}$. We can represent the field observed at $\udot{A}$ as $f(\dvec{p}^A,t)$, where position is represented by $\dvec{p}^A=[x_A,y_A,z_A]^T\in\mathbb{R}^3$ and time by $t\in\mathbb{R}$,

@@

%
\begin{align}
f_\textrm{plane}(\dvec{p}^A,t)  &= A e^{j(\omega t - \dvec{k}^T \dvec{p}^A)} \\
f_\textrm{spheric}(\dvec{p}^A,t) &= \frac{A}{\lVert \dvec{p}^A-\dvec{p}_{AS}^A \rVert} e^{j(\omega t - \dvec{k}^T \lVert \dvec{p}^A-\dvec{p}_{AS}^A \rVert)}
\end{align}
%
% \hyperref[IV_methods]{\chapterref{IV_methods}.\sectionref{IV_methods}}
% \renewcommand{\thesection}{\thechapter.\arabic{section}}

@@

Consider a sonar imaging scenario where an encoded signal is transmitted to highlight a surface of interest, and assume that the backscattered wavefield is properly sampled using a uniform linear array of receivers. After stripping the carrier frequency by heterodyning and the signal signature by  matched filtering (Fig. \ref{I_beamforming}), an image can then be formed by coherently combining the receiver outputs to focus at one angle (or point) at the time. The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

Suppose we record a wavefield with $M$ sensors arranged in a uniformly spaced linear array (ULA) and attach a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length). Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$. We simply notation by making the dependence on $\theta$ implicit from now on.

The beamformer output $z[n]\in\mathbb{C}$ is by definition the weighted sum of all the delayed data samples:
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{2_eq_z}
\end{align}
%
where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}. However, there exists a class of adaptive beamformers that target this limitation by adapting the weights to the impinging wavefield. The MVDR method is one such technique.

@@

\subsubsection{Complexity and implementation}

If (\ref{2_eq_spatialR}) were to be implemented directly, we note that the estimation step would have a complexity of O($N_K N_L L^2$). By comparison, performing a Gauss Jordan inversion is generally of O($L^3$), which implies that whenever $N_K\,N_L = N_K(M-L+1) \gg L$ the estimation step will dominate. This tends to be true in active systems because we often want to set $L<M/2$ and $N_K\ge>3$ for the MVDR to be robust enough to yield the expected results.

A few properties let us improve these matters notably. First, we may infer from (\ref{2_eq_spatialR}) that:
%
\begin{itemize}
\item The subarray covariance matrices $R_l[n] = \x_l[n]\x_l^H[n]$ are Hermitian semi-definite, which means that we can get away with computing only one of their triagonals.
\item When summing $R_l[n]$ most of the multiplications need only be computed only once, and most sums can be updated iteratively.
\end{itemize}
%
Second, each of the diagonals of $R_l[n]\ \forall\ l$ can be computed independently, which is good news for our GPU design.

@@

\section{Implementation}\label{implementation}

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=\linewidth]{gfx/mvdr_implementation.svg}%
\caption{MVDR implementation.}\label{2_eq_mvdr_implementation}
\end{figure}

Our GPU implementation consists of three main kernels, one for each of the computing steps illustrated in Figure \ref{2_eq_mvdr_implementation}. In the first step we estimate $\eR$, in the second we solve the linear equation $\eRi\1$, and finally we perform some substitutions to obtain the beamformer output, $z$. All pixels are processed by one step before we move on to the next step. Our efforts here were mainly focused on estimating $\eR$, because a highly optimised batch linear equation solver is already available from Nvidia, and computing the beamformer output is a rather simple task.

When computing $\eR$ we we assign $L$ threads per pixel to traverse the diagonals of $\eR$. On the top row a full computation is carried out, then that entire row is saved back to global memory following a collective access pattern that maximizes global memory bandwidth. For subsequent rows the threads move along the diagonals while performing iterative summations; the result from the previous element on the diagonal is updated by adding and removing the correlation coefficients that enters and exits the sum, respectively. To minimize memory consumption, we compute the coefficients again every time we need them. When a thread has finished up a diagonal, we have them wrap around to compute one of the diagonals in the lower triangular of $\eR$. Since $\eR$ is conjugate symmetric, the values in the leftmost column is obtained by a complex conjugate copy of the relevant value in the first row. This procedure balances the load evenly on all threads, is almost completely free of arithmetic redundancy, and consumes very little memory.

The implementation of beamspace MVDR is very similar, but the data access pattern is different. In the time domain subsequent subarrays overlap on all but one element in memory, but the Fourier transformed data does not overlap at all. Such a seemingly innocent detail can make all the difference in a GPU design, because designing to make efficient use of memory is key to obtaining good performance.

>>>>

Waves are made of endless tiny energy transactions, a dance of particles effecting each other in search of equilibrium. They are caused by disturbances in the forces of nature. The drop of a rock into calm water or rapid displacement of still air cause mechanical waves, while synchronized oscillations of electric and magnetic fields cause electromagnetic waves. Regardless of whether the waves are mechanical or electromagnetic, move perpendicular or parallel to the axis of propagation, what medium they propagate through or the means by which they do so, they are---perhaps surprisingly---all governed by a partial differential equation known as the \emph{wave equation}:
%
\begin{align}
\Delta f = 
\frac{\delta^2 f}{\delta x^2} + \frac{\delta^2 f}{\delta y^2} + \frac{\delta^2 f}{\delta z^2} = \frac{1}{c}\frac{\delta^2 f}{\delta t^2},
\end{align}
%
where $\Delta$ is the Laplace operator, $f$ is field strength and $c\in\mathbb{R}^+$ is wave propagation speed. If $f$ represents the displacement of a vibrating string, the equation states that the force pulling the string towards equilibrium (right side) is proportional to the string's curvature (left side). In acoustics $f$ represents sound pressure, but since the mathematics are all the same we will simply think of $f$ as a signal: a unit-less scalar function with wave characteristics.

A few solutions to the wave equation is of particular interest, that of a plane and spherical wave
%
\begin{align}
f_\textrm{plane}(\dvec{p}^A,t)  &= A e^{j(\omega t - \vec{k} \cdot \vec{p}_{SO})} \\
f_\textrm{spheric}(\dvec{p}^A,t) &= \frac{A}{\lVert \dvec{p}_{SO}^S \rVert} e^{j(\omega t - \dvec{k}^T \lVert \dvec{p}_{SO}^A \rVert)}
\end{align}
%

One is that of plane wave and the other a  A signal source with the position $\udot{S}$ . We can represent the field observed at $\udot{A}$ as $f(\dvec{p}^A,t)$, where position is represented by $\dvec{p}^A=[x_A,y_A,z_A]^T\in\mathbb{R}^3$ and time by $t\in\mathbb{R}$,

@@

%

%
% \hyperref[IV_methods]{\chapterref{IV_methods}.\sectionref{IV_methods}}
% \renewcommand{\thesection}{\thechapter.\arabic{section}}

% Suppose an array of sensors has a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ attached to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length).

@@

Consider a sonar imaging scenario where an encoded signal is transmitted to highlight a surface of interest, and assume that the backscattered wavefield is properly sampled using a uniform linear array of receivers. After stripping the carrier frequency by heterodyning and the signal signature by  matched filtering (Fig. \ref{I_beamforming}), an image is formed by coherently combining the receiver outputs to focus at one angle (or point) at a time. The principle of adjusting the array's focus in range and direction is commonly referred to as beamforming, and involves assigning suitable delays and weights to the array's channels.

@@

Imagine we record a wavefield with $M$ sensors arranged in a uniformly spaced linear array (ULA) and attach a reference frame $\ubar{A} \triangleq (\udot{A},\uvec{A})$ to it, where $\udot{A}\in\mathbb{A}$ is its absolute position and $\uvec{A}\triangleq\big(\hat{x}_{A},\hat{y}_{A},\hat{z}_{A}\big)\in\mathbb{V}^3$ its absolute orientation (where $\hat{\cdot}$ denotes unit length). Assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n-\Delta_m]\in\mathbb{C}$. We simplify notation by making the dependence on $\theta$ implicit from now on.

The beamformer output $z[n]\in\mathbb{C}$ is by definition the weighted sum of all the delayed data samples:
%
\begin{align}
z[n] = \dvec{w}^H[n]\dvec{x}[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{2_eq_z}
\end{align}
%
where $w_m\in\mathbb{C}$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. A large variety of weighting functions exist for trading lateral resolution for improved noise suppression (contrast), but a compromise between the two is unavoidable~\cite{Harris1978}. However, there exists a class of adaptive beamformers that target this limitation by adapting the weights to the impinging wavefield. The MVDR method is one such technique.

@@

@@
